---
title: "Assignment 2: Coding Basics"
author: "David Robinson"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. 
#Note that this sequence stops at 28 instead of 30 because adding 3 to 28 would result in 31.
#QUESTION -- Is this acceptable?
seq(1,30,3)

#2. 
sequence_1To30 <- seq(1,30,3)
mean_sequence1To30 <- mean(sequence_1To30)
median_sequence1To30 <- median(sequence_1To30)

#3. 
#Asking R to determine whether the mean is greater than the median via Boolean logic.
mean_sequence1To30>median_sequence1To30
#Asking R to determine whether the mean is greater than the median via Boolean logic and and if / else statement.
ifelse(mean_sequence1To30>median_sequence1To30,"The mean is greater than the median.","The mean is not greater than the median.")

```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
#5 and #6
#This vector is a list of string (character) values.
StudentNames <- c("Leonardo", "Raphael","Michelangelo", "Donatello")

#This vector is a list of integer values.
TestScores <- c(100, 75, 50, 25)

#This vector is a list of Boolean (logical) values.
TestScoresPassing <- c(TRUE, TRUE, TRUE, FALSE)

#7 and #8
#QUESTION -- Are the columns of the data frame not already labeled with the informative titles of their composite columns?
#QUESTION -- Are the commas after the numbers in the row labels acceptable?
is.data.frame(df_StudentNames)
df_StudentNames <- as.data.frame(StudentNames)
df_StudentNamesAndTestScores <- cbind(StudentNames,TestScores,TestScoresPassing)
df_StudentNamesAndTestScores

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer:
#9
Matrices can only contain a single class of data, while data frames can consist of many different classes of data (https://carpentries-incubator.github.io/lc-litsearchr/07-data-frames-matrices-and-lists/index.html#:~:text=The%20main%20difference%20is%20that,grid%20of%20rows%20and%20columns). 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 
```{r if / else function for vector of test scores}

PassingScore <- 50
# Pass <- "TRUE"
# Fail <- "FALSE"
Pass <- TRUE
Fail <- FALSE
#QUESTION -- what is the difference between storing these as Boolean values versus strings as it relates to later printing them?
#QUESTION -- can the entire thing be wrapped in a print statement and call it good?

#QUESTION -- does the input here need to be the entire vector within a single function (seems to be how the question is written) or should it be a series of point-values within a vector across multiple functions?

print(ifelse(TestScores[1:4]>=PassingScore,Pass,Fail))

#QUESTION -- what causes this to "print twice"? I think one is showing the returned value and one is printing, but I'm not sure how to "only print."


  
```


11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}

#11
#QUESTION -- What is the distinction between this question and the question above? Is the suggesiton to exclude "print" here? Or was it to extend from a single point in the vector to the entire thing? Or to evaluate the entire vector as an average or something?

print(ifelse(TestScores[1:4]>=PassingScore,Pass,Fail))

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:
QUESTION -- Does this make sense? Could either option really be used?
I used the 'ifelse' option. I think either option could work, but 'ifelse' was more familiar to me based on my experience in Excel and Python.
