---
title: "Assignment 10: Data Scraping"
author: "David Robinson"
output: pdf_document
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 

#install.packages("tidyverse")
library(tidyverse)
#install.packages("rvest")
library(rvest)
#install.packages("here")
library(here); here()


```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 

webpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022')
webpage

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 

the_water_system_name <- webpage %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text()
the_water_system_name

the_PWSID <- webpage %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text()
the_PWSID

the_ownership <- webpage %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text()
the_ownership

the_MGD <- webpage %>%
  html_nodes("th~ td+ td") %>%
  html_text()
the_MGD

#QUESTION -- does the order of the months need to be modified later?

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data}
#4 

the_months <- c("Jan", "May", "Sept", "Feb", "Jun", "Oct", "Mar", "Jul", "Nov", "Apr", "Aug", "Dec")
the_years <- c("2022", "2022", "2022", "2022", "2022", "2022", "2022", "2022", "2022", "2022", "2022", "2022")

df_withdrawals <- data.frame("Month" = the_months,
                             "Year" = the_years,
                             "Avg_Withdrawals_mgd" = as.numeric(the_MGD))
df_withdrawals

df_withdrawals_full <- df_withdrawals %>%
  mutate(Water_System_Name = !!the_water_system_name,
         PWSID = !!the_PWSID,
         Ownership = !!the_ownership,
         Date = my(paste(Month,"-",Year)))

df_withdrawals_full

#5 

ggplot(df_withdrawals_full,aes(x=Date,y=Avg_Withdrawals_mgd)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2022 Water usage data for",the_ownership),
       subtitle = the_water_system_name,
       y="Withdrawal (mgd)",
       x="Date")

#QUESTION -- are these months in the right order or does something need to happen to sort them?

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. 

the_base_url <- 'https://www.ncwater.org/WUDC/app/LWSP/report'
'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022'
the_PWSID_var <- '03-32-010'
the_year_var <- 2022
the_scrape_url <- paste0(the_base_url, '.php?pwsid=', the_PWSID_var, '&year=', the_year_var)
print(the_scrape_url)

#QUESTION -- is this saying now to copy / paste most of 2. - 5. to complete the function, or is the above sufficient?

#Retrieve the website contents
webpage_var <- read_html(the_scrape_url)
webpage_var

#Set the element address variables (determined in the previous step)
the_water_system_name_tag <- webpage_var %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text()

the_PWSID_tag <- webpage_var %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text()

the_ownership_tag <- webpage_var %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text()

the_MGD_tag <- webpage_var %>%
  html_nodes("th~ td+ td") %>%
  html_text()

#Scrape the data items
the_water_system_name_var <- webpage_var %>%
  html_nodes(the_water_system_name_tag) %>%
  html_text()

the_PWSID_var <- webpage_var %>%
  html_nodes(the_PWSID_tag) %>%
  html_text()

the_ownership_var <- webpage_var %>%
  html_nodes(the_ownership_tag) %>%
  html_text()

the_MGD_var <- webpage_var %>%
  html_nodes(the_MGD_tag) %>%
  html_text()

#Construct a dataframe from the scraped data
df_withdrawals_var <- data.frame("Month" = the_months,
                             "Year" = the_years,
                             "Avg_Withdrawals_mgd" = as.numeric(the_MGD_var)) %>%
df_withdrawals_var_full <- df_withdrawals_var %>%
  mutate(Water_System_Name = !!the_water_system_name_var,
         PWSID = !!the_PWSID_var,
         Ownership = !!the_ownership_var,
         Date = my(paste(Month,"-",Year)))

#Plot
ggplot(df_withdrawals_var,aes(x=Date,y=Avg_Withdrawals_mgd)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2022 Water usage data for",the_ownership_var),
       subtitle = the_water_system_name_var,
       y="Withdrawal (mgd)",
       x="Date")

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 

#QUESTION -- look back into function syntax from A02 or A03


```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 




```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 




```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer:
>
