---
title: "Assignment 3: Data Exploration"
author: "David Robinson"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r R session set up,  message = FALSE}

#Check working directory
getwd()

#Load necessary packages
#install.packages("tidyverse")
#QUESTION -- I don't think the below line is necessary?
#library(tidyverse)
#install.packages("lubridate")
#QUESTION -- I don't think the below line is necessary?
#library(lubridate)

#Upload the two datasets in question
Neonics <- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv",
                    stringsAsFactors = TRUE)
Litter <- read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv",
                   stringsAsFactors = TRUE)

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer:
The very existence and use of insecticides (e.g., neonicotinoids) is driven by insects who are percieved as "pests" as they interact with (perhaps by eating) crops in agriculture. One reason why we might be interested in the ecotoxicology of neonicotinoids is that they could be affecting insects other than those considered pests in a given agricultural scenario. These other, non-pest
insects may even
be beneficial to human needs by functioning as natural predators for pests or as pollinators for crops.
Source: https://www.pnas.org/doi/10.1073/pnas.2017221117

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer:
Quoting directly from the below source: "Woody debris is an important part of forest and stream ecosystems because it has a role in carbon budgets and nutrient cycling, is a source of energy for aquatic ecosystems, provides habitat for terrestrial and aquatic organisms, and contributes to structure and roughness, thereby influencing water flows and sediment transport (Harmon and others 1986)."
Source: https://www.fs.usda.gov/research/treesearch/20001#:~:text=Woody%20debris%20is%20an%20important,influencing%20water%20flows%20and%20sediment

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Litter is defined as material that is dropped from the forest canopy and has
 a butt end diameter <2cm and a length <50 cm; this material is collected in elevated PVC traps where each is a 0.5m^2 square with mesh ‘basket’ elevated
~80cm above the ground. Fine wood debris is defined as material that is dropped from the forest canopy and has a butt end diameter <2cm and a length >50 cm;
this material is collected in ground traps as longer material is not reliably collected by the elevated traps. Ground traps are 3 m x 0.5 m rectangular areas.
2.Litter and fine woody debris sampling is executed at terrestrial NEON sites
that contain woody vegetation >2m tall. Along with most of NEON’s plant productivity measurements, sampling for this product occurs only in tower
plots. Locations of tower plots are selected randomly within the 90% flux footprint of the primary and secondary airsheds (and additional areas in close proximity to the airshed, as necessary to accommodate sufficient spacing between plots).
3. Ground traps are sampled once per year. Target sampling frequency for
elevated traps varies by vegetation present at the site, with frequent sampling (1x every 2weeks) in deciduous forest sites during senescence, and infrequent
year-round sampling (1x every 1-2 months) at evergreen sites.


## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r dimensions of dataset}

dim(Neonics) #Neonics has 4623 rows and 30 columns

```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r summary of dataset}
summary(Neonics$Effect) #Produce summary of Effect column within Neonics data

```

> Answer:
"Mortality" and "Population" are the most common effects that are studied. These effects are of interest given that they could be indicators for the potential of neonicotinoids to harm beneficial insects and negatively impact agricultural practices.


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r most common species}
sort(summary(Neonics$Species.Common.Name)) #Produce sorted summary of Species Common Name with Neonics data

```

> Answer:
The six most commonly studied species (other than "(Other)" which is not a
species name) are Honey Bee, Parasitic Wasp, Buff Tailed Bumblebee, Carniolan Honey Bee, Bumble Bee, and Italian Honeybee. These species are all bees or wasps and all
have a role in the pollinator ecosystem, which is of great importance for maximizing crop yield.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r class of conc.1..author}
class(Neonics$Conc.1..Author) #Check the class of 'Conc.1.Author'

```

> Answer:
The class is "factor." Storing these concentrations as values allows them to be interpreted in a manner like that of categorical variables versus solely as numbers.
QUESTION -- is this correct? Is this covered in the videos somewhere? Note that some also have "/" character. The concentrations are also in different units...
so perhaps they shouldn't be added together, averaged, etc. to prevent confusion.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r generate a plot of the number of studies conducted by publication year}
#class(Neonics$Publication.Year)
#Neonics$Publication.Year <- as.Date(Neonics$Publication.Year, format = )
#QUESTION -- is this necessary? If yes, what is the proper format?
#QUESTION -- is there any constriction here for the number of bins?

#Generate a plot of the number of studies conducted by publication year with 30 bins
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year), bins = 30)

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r number of studies conducted by publication year with color}

#Generate a plot of the number of studies conducted by publication year with 30 bins and different test locations displayed as different colors
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year, color = Test.Location), bins = 30)

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer:
"Lab" and "Field natural" are  the two most common test locations. While these
are clearly the two most common test locations overall since 1990, there is a brief moment in ~2003 in which "Field artificial" appears to be more common than "Field natural." Lab and Field natural do differ over time -- e.g.,
Field natural is more common than Lab from ~1993 - 2000, but then Lab is more common until ~2008, then briefly back to Field natural. After 2010, Lab is most common.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r create bar graph of Endpoint counts}

#Generate a bar graph of the endpoint counts
ggplot(Neonics, aes(x = Endpoint)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

> Answer:
The two most common points are "NOEL" and "LOEL".
Per ECOTOX_CodeAppendix...
"NOEL" is defined as "No-observable-effect-level: highest dose (concentration) producing effects not significantly different from responses of controls
according to author's reported statistical test."
"LEOL" is defined as "Lowest-observable-effect-level: lowest dose
(concentration) producing effects that were significantly different
(as reported by authors) from responses of controls."

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r determine class of collectDate}

#class(Litter$collectDate) #The class of collectDate was initially a factor
Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d") #The class of collectDate is now a date
class(Litter$collectDate) #The class of collectDate is confirmed as a date
unique(Litter$collectDate) #Litter was sampled on the 2nd and the 30th of August 2018


```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r determine how many plots were sampled at Niwot Ridge}

unique(Litter$plotID) #Determine how many unique plots were sampled
summary(Litter$plotID) #Produce summary of how many times each unique plot was sampled 

```

> Answer:
12 plots were sampled at Niwot Ridge. The 'unique' function provides a list of the unique values within a set and tells us how many levels there are. The 'summary' function provides a count of the frequency of each of those unique values within a set.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r create bar graph of functionalGroup counts}

#Generate a bar graph of the functional group counts
ggplot(Litter, aes(x = functionalGroup)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r create boxplot and violin plot of dryMass by functionalGroup}

#Generate a boxplot of dry mass by functional group
ggplot(Litter) +
  geom_boxplot(aes(x = dryMass, y = functionalGroup))

#Generate a violin plot of dry mass by functional group
ggplot(Litter) +
  geom_violin(aes(x = dryMass, y = functionalGroup))

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:
A box plot allows us to look at a range of values. A violion plot allows us to
look at both a range of values and a distribution of values within that range.
In this case, a violin plot allows us to see very little, which suggests that
the distribution within the range of values is not concentrated / dense in any specific area(s). Given this feature of the data, the boxplot allows us to more clearly see the range of the data standalone.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:
Per the boxplot, Needles and Mixed litter tend to have the highest biomass.
Their median biomass values are higher than the maximum value of most other
litter types (except for Twigs/branches, which has a maximum biomass value that
is greater than the median value for Mixed).
