---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "David Robinson"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1

#Load necessary packages

#install.packages("tidyverse")
library(tidyverse)
#install.packages("lubridate")
library(lubridate)
#install.packages("here")
library(here)
#install.packages("cowplot")
library(cowplot)
#install.packages("ggthemes")
library(ggthemes)
#install.packages("agricolae")
library(agricolae)

#Check working directory
here()

#Import relevant data files
Lake <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), stringsAsFactors = TRUE)
Lake$sampledate <- as.Date(Lake$sampledate,format = "%m/%d/%Y")

#2

#QUESTION -- do this later -- could just recycle theme from A05
#See 13:56 from video 1


```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature recorded during July does not change with depth across all lakes.
Ha: Mean lake temperature recorded during July does change with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

#Wrangle per the requested criteria
Lake_wrangled <- Lake %>%
  filter(month(sampledate) == 7) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  na.omit()

#5

#Visualize the relationship
ggplot(Lake_wrangled,
       aes(
         y = temperature_C,
         x = depth
       )) +
  geom_point() +
  ylim(0,35) +
  geom_smooth(method = lm) +
  labs(y = "Temperature (C)",
       x = "Depth")


```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: 
Temperature decreases as depth increases. The distribution suggests a weak, negative correlation within a roughly linear trend.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7

#Perform linear regression
Lake.regression <- lm(data = Lake_wrangled,
                     temperature_C ~ depth)

summary(Lake.regression)


```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:
Based on the slope, we see that the temperature is decreasing by 1.94621 degrees per meter of depth.
Based on the R-squared value of 0.7387, around 73.8% of the variability in temperature is explained by changes in depth.
Degrees of freedom = 9726
Statistical significance = Given the low p-value of 2e-16 (less than 0.05), we can reject the null hypothesis (H0: Mean lake temperature recorded during July does not change with depth across all lakes). Thus, the mean lake temperature recorded during July does change with depth across all lakes.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

#Run AIC and step
Lake.regression.AIC <- lm(data = Lake_wrangled, temperature_C ~ year4 + daynum + depth)

step(Lake.regression.AIC)


#10

#Summarize multiple regression
summary(Lake.regression.AIC)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 
The final set is to "remove none" which leaves year4, daynum, and depth.
Based on the R-squared value of 0.7412, around 74.1% of the observed variance in this model is explained. Yes, this is a slight improvement over using only depth as the explanatory variable which had an R-squared value of 0.7387.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

#ANOVA model
Lake_wrangled_ANOVA <- aov(data = Lake_wrangled, temperature_C ~ lakename)
summary(Lake_wrangled_ANOVA)


#Linear model
Lake_wrangled_ANOVA_2 <- lm(data = Lake_wrangled, temperature_C ~ lakename)
summary(Lake_wrangled_ANOVA_2)



```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: 
Based on the p-value of less than 2e-16 (less than .05), we reject the null hypothesis (H0 = the mean temperature is the same across all the lakes). Thus, the mean temperatures are not the same across all the lakes.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14

#Create graph and make it pretty (via axis labels)
ggplot(Lake_wrangled,
       aes(
         y = temperature_C,
         x = depth,
         color = lakename
       )) +
  geom_point(alpha = 0.5) +
  ylim(0,35) +
  geom_smooth(method = lm, se = FALSE) +
  labs(y = "Temperature (C)",
       x = "Depth",
       color = "Lake Name")


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15

#Tukey HSD test
print(HSD.test(Lake_wrangled_ANOVA, "lakename", group = T))


```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:
Ward Lake and Paul Lake have the same mean temperature, statistically speaking, as Peter Lake.
No lake has a mean temperature that is statistically distinct from all other lakes (based on the output of the Tukey HSD in which no lake has a distinct group identifier).


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: 
A two-sample T-test.


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
#18

#Wrangle the data
Lake_wrangled_18 <- Lake_wrangled %>%
  filter(lakename == "Crampton Lake" |
         lakename == "Ward Lake")

#Run the two-sample T-test
t.test(Lake_wrangled_18$temperature_C ~Lake_wrangled_18$lakename)


```

>Answer: 
Because the p-value is greater than 0.05, we cannot reject the null hypothesis (H0 = the mean temperature is the same between Crampton Lake and Ward Lake). Thus, the mean temperatures between the two lakes are the same. Yes, this does match the answer for #15 / #16 in which Crampton Lake and Ward Lake were both in group b.


