---
title: "Assignment 5: Data Visualization"
author: "David Robinson"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 

#Load necessary packages

#install.packages("tidyverse")
library(tidyverse)
#install.packages("lubridate")
library(lubridate)
#install.packages("here")
library(here)
#install.packages("cowplot")
library(cowplot)
#install.packages("ggthemes")
library(ggthemes)

#2 

#Read in the data sets in question

Nutrients <- read.csv(file = here('Data','Processed_KEY', 'NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv'), stringsAsFactors = T)

Litter <- read.csv(file = here('Data','Processed_KEY', 'NEON_NIWO_Litter_mass_trap_Processed.csv'), stringsAsFactors = T)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3

#Customized axis label size
#Customized axis label color
#Customized label position

mytheme <- theme_base() +
  theme(
    line = element_line(
      color = 'blue',
      linewidth = 3
    ),
    legend.background = element_rect(
      color = 'black',
      fill = 'red'
    ),
    legend.title = element_text(
      color = 'orange'
    )
  )

#QUESTION -- figure out what to do with the below

    #line =              element_line(),
    #rect =              element_rect(),
    #text =              element_text(),
    
    # Modified inheritance structure of text element
    #plot.title =        element_text(),
    #axis.title.x =      element_blank(),
    #axis.title.y =      element_blank(),
    #axis.text =         element_text(),
    
    # Modified inheritance structure of line element
    #axis.ticks =        element_line(),
    #panel.grid.major =  element_line(),
    #panel.grid.minor =  element_blank(),
    
    # Modified inheritance structure of rect element
    #plot.background =   element_rect(),
    #panel.background =  element_rect(),
    #legend.key =        element_rect(),
    
    # Modifiying legend.position
    #legend.position = 'right',
    
    #complete = TRUE
    #)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 

#QUESTION -- how to apply 'mytheme' here?
#QUESTION -- what does 'separate aesthetics' mean here? Is color sufficient? Also need to change the shape of the points?

ggplot(Nutrients,
       aes(
         y = po4,
         x = tp_ug,
         color = lakename
       )
       ) +
  geom_point() +
  #QUESTION -- this doesn't work
  #scale_shape_manual(values = c(15, 17)) +
  xlim(0, 150) +
  ylim(0, 150) +
  geom_smooth(method= lm, color = "black")


```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip:
* Recall the discussion on factors in the previous section as it may be helpful here. 
* R has a built-in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 

boxplot_temp <- ggplot(Nutrients,
       aes(x = month,
           y = temperature_C
           )) +
         geom_boxplot(aes(color = lakename))

boxplot_TP <- ggplot(Nutrients,
       aes(x = month,
           y = tp_ug
           )) +
         geom_boxplot(aes(color = lakename))

boxplot_TN <- ggplot(Nutrients,
       aes(x = month,
           y = tn_ug
           )) +
         geom_boxplot(aes(color = lakename))

#QUESTION -- the below counts as a 'cowplot that combines the three graphs', correct?
#QUESTION -- did not use factors per tip
#QUESTION -- did not use 'month.abb' per tip -- is this to replace month numbers with names?
#QUESTION -- is the replication of month along the bottom an issue?
#QUESTION -- axes are aligned but number scales are not the same -- adjust?

plot_grid(boxplot_temp + ylab(NULL) + theme(legend.position = 'none'),
          boxplot_TP + ylab(NULL) + theme(legend.position = 'none'),
          boxplot_TN + ylab(NULL) + theme(legend.position = 'none'),
          get_legend(boxplot_temp), #+ theme(legend.position = 'bottom'),
          nrow = 1, 
          align = 'vh', 
          labels = c('Temp', 'TP', 'TN'), 
          axis = 'b')

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 
x

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6




#7




```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:
x